Follow-up Samples

### 2017-02-17 submission/manuscript_outline_20161024.Rmd

**Background:** Colorectal cancer (CRC) continues to be a worldwide health problem with previous research suggesting that a link may exist between the fecal bacterial microbiome and CRC.  The overall objective of our study was to test the hypothesis that changes in the bacterial microbiome occur after surgery in patients with lesions (i.e. adenoma or carcinoma).  Specifically, we wanted to identify what within the community was different before and after surgical removal of said lesion.

**Results:** The bacterial microbiome in pre and post surgery samples of individuals with adenoma were more similar to each other than those with carcinoma (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.05, paste("<"), paste(">"))` 0.05).  There was no difference in the relative abundance of any OTU between the pre and post surgery samples (P-value `r ifelse(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']) < 0.05, paste("<"), paste(">"))` 0.05).  A model with a total of `r length(lesion_top10_varsMDA$variable)` variables was able to classify lesion (AUC = `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) while a model built to classify pre surgery samples had `r length(IF_top10_varsMDA$variable)` variables (AUC = `r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)`).  The post surgery sample for both models had a decrease in the positive probability for either lesion or pre surgery sample (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).  In total there were `r length(common_vars_summary$otu)` OTUs that were common to both models and the majority of these classified to commensal bacteria (e.g. *`r filter(common_vars_summary, otu == "Otu000028")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000018")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000009")[, "lowest_ID"]`*, and *`r filter(common_vars_summary, otu == "Otu000005")[, "lowest_ID"]`*).

**Conclusions:** Our data supports the hypothesis that there are differences in the bacterial microbiome between pre and post surgical samples.  With individuals with carcinoma having more drastic differences to the overall community then those with adenoma.  Changes to commensal bacteria were some of the most important variables for model classification, suggesting that these bacteria may be central to initial polyp formation and transition to carcinoma.

@@

Recently, there has been promising work on the bacterial microbiome and it's ability to be able to complement existing screening methods such as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014]. There has also been research into how this microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].  A few studies have also shown how this microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the pathogenesis of CRC.  These studies have helped to provide a tantalizing link between the bacterial microbiome and CRC. Although these studies suggest that the bacterial microbiome might change after treatment there remains limited information on the bacterial microbiome before and after surgical removal of lesion (adenoma or carcinoma) and whether the community changes at all.

In this study we tested the hypothesis that the bacterial microbiome changes between pre (initial) and post (follow up) samples after surgery for individuals with a lesion (adenoma or carcinoma).  Our analysis included both alpha and beta diversity analysis along with investigation of individual operational taxonomic units (OTUs).  We also utilized Random Forest models and observed how these models, as well as specific OTUs within them, performed pre and post surgery.  We also investigated the models for similar important OTUs to identify the crucial OTUs for not only classifying initial and follow up samples but also lesion or normal.

@@

***Cancer Associated Bacteria:***
Previous literature has suggested that a number of oral microbes may be important in CRC pathogenesis [@flynn_metabolic_2016].  So we next examined whether there were differences in previously well described carcinoma associated OTUs.  These included the OTUs that aligned with *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).  There was a difference in relative abundance in initial and follow up samples for lesion and carcinoma for *Parvimonas micra* (P-value `r ifelse(main_crc_bugs_summary["parv", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05), and *Porphyromonas asaccharolytica* (P-value `r ifelse(main_crc_bugs_summary["porp", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05). In contrast, there was no difference in relative abundance in any of these OTUs for those with adenoma [Figure 2]. We also observed that only a small percentage of those with adenoma or carcinoma were positive or had an appreciable relative abundance of any of these respective OTUs [Figure 2].

***Full and Reduced Model:***
We next wanted to identify if there were any common bacterial microbiome changes in individuals with adenoma or carcinoma.  In order to investigate this we created two different models: one to classify lesion versus normal and one to classify pre (initial) versus post (follow up) samples based on the bacterial community and FIT measurements.  The lesion model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` to `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` while the initial sample model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.  By identifying the most important variables for each respective model and then reducing them to only these factors we were able to increase the AUC in the lesion model (`r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) and initial sample model (`r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)`).        
The test set AUC range for the full and reduced lesion model were similar to that reported for the training set AUC ranges and the ROC curve ranges overlap with each other [Figure 3a].  The ROC curve for the final lesion model used was within the range of the reduced lesion model [Figure 3a].  Interestingly, the test set AUC range for the initial sample model performed much better then the training set AUCs.  Both the full and reduced initial sample models overlapped with each other [Figure 3b] but there was a marked decrease in the ROC curve for the final before sample model used.

***Common OTUs to both Models:***
The reduced models were built based on the most important variables to either classification model.  For the lesion model there were a total of `r length(lesion_top10_varsMDA$variable)` variables [Figure S2] whereas for the initial sample model there were a total of `r length(IF_top10_varsMDA$variable)` variables [Figure S3].  For both models FIT measurement resulted in one of the largest decreases in MDA [Figure S2a & S3a].  When we compared the two different reduced models with each other there were a total of `r length(common_vars_summary$otu)` common OTUs.  Some of the most common taxonomic identifications belonged to *`r filter(common_vars_summary, otu == "Otu000028")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000018")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000009")[, "lowest_ID"]`*, and *`r filter(common_vars_summary, otu == "Otu000005")[, "lowest_ID"]`*.  The vast majority of these OTUs had classifications to bacteria typically thought of as commensal [Table S2].

@@

## Discussion
From our results there were large observed differences in the bacterial microbiome between pre and post surgery samples based on whether the individual had an adenoma or carcinoma.  There were much larger differences between initial and follow up samples based on the thetayc distance metric and in fecal blood as measured by FIT for individuals with carcinoma versus adenoma [Figure 1].  However, there were no differences between initial and follow up samples for Shannon Diversity, observed OTUs, or evenness regardless of whether the individual had an adenoma or carcinoma [Table S1].  There was also no differences in relative abundance of any specific OTU for lesion, adenoma only, or carcinoma only [Figure s1].  Although there was a detectable change in the carcinoma group.  It should be noted that whether this change was towards what wouldbe be expected for normal controls is not known.

@@

***Analysis Overview:***
We first wanted to test if there were any differences based on whether the individual had an adenoma or carcinoma.  This was done by testing initial and follow up samples for differences in alpha and beta diversity, testing differences in FIT between initial and follow ups, testing all OTUs that were used in either built model, and investigating the relative abundance of specific previously associated CRC bacteria (*Fusobacterium nucleatum*, *Parvimonas micra*, *Peptostreptococcus assacharolytica*, and *Porphyromonas stomatis*) based on adenoma and carcinoma.  From here the lesion model was then tested for accuracy in prediction and whether it reduced the positive probability of lesion after surgery.  The most important OTUs for this were used to build a reduced model and it was assessed for similarity to the original model.  We then used the initial sample model to assess whether it could classify samples better then the lesion model.  The most important OTUs were then identified from this model and used to create a reduced feature initial sample model.  This reduced feature model, as was done with the lesion model, was compared to the full model for loss of accuracy.  A list of common OTUs were found for the two different models used.  Finally, both lesion and initial sample models initial positive probabilites were tested against the follow up positive probabilities.

@@

**Figure 4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Initial Sample Reduced Variable Models**  A) Lesion positive probability change from initial to follow up sample in those with carcinoma   B) Initial positive probability change from initial to follow up sample in those with carcinoma   C) Lesion positive probability change from initial to follow up sample if those with adenoma or Screen Relevent Neoplasia (SRN).    D) Initial positive probability change from initial to follow up sample in those with adenoma or SRN.  The dotted line represents the threshold used to make the decision of whether a sample was positive or not.

@@

**Figure S4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Before Sample Full Variable Models** A) Lesion positive probability change from initial to follow up sample in those with carcinoma   B) Initial positive probability change from initial to follow up sample in those with carcinoma   C) Lesion positive probability change from initial to follow up sample in those with adenoma or Screen Relevent Neoplasia (SRN).    D) Initial positive probability change from initial to follow up sample in those with adenoma or SRN.  The dotted line represents the threshold used to make the decision of whether a sample was positive or not.

>>>>

**Background:** Colorectal cancer (CRC) continues to be a worldwide health problem with previous research suggesting that a link may exist between the fecal bacterial microbiome and CRC.  The overall objective of our study was to test the hypothesis that changes in the bacterial microbiome occur after lesion (i.e. adenoma or carcinoma) removal.  Specifically, we wanted to identify what within the community was different before and after removal of said lesion.

**Results:** The bacterial microbiome in pre and post surgery samples for the adenoma group were more similar to each other than the carcinoma group (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.05, paste("<"), paste(">"))` 0.05).  There was no difference for either the adenoma or carcinoma group in the relative abundance of any OTU between the pre and post lesion removal (P-value `r ifelse(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']) < 0.05, paste("<"), paste(">"))` 0.05).  A model built to classify model built to classify lesion had an AUC range of `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)` while a model built to classify initial versus follow up samples had an AUC range of `r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)`.  The post removal sample for both models had a decrease in the positive probability for either lesion or initial sample (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).  The lesion model used a total of `r length(lesion_top10_varsMDA$variable)` variables while the initial sample model used a total of `r length(IF_top10_varsMDA$variable)` variables.  A total of `r length(common_vars_summary$otu)` OTUs were common to both models with the majority of these classifying to commensal bacteria (e.g. *`r filter(common_vars_summary, otu == "Otu000028")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000018")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000009")[, "lowest_ID"]`*, and *`r filter(common_vars_summary, otu == "Otu000005")[, "lowest_ID"]`*).

**Conclusions:** Our data supports the hypothesis that there are differences in the bacterial microbiome between pre and post lesion removal samples.  With individuals with carcinoma having more drastic differences to the overall community then those with adenoma.  Changes to commensal bacteria were some of the most important variables for model classification, suggesting that these bacteria may be central to initial polyp formation and transition to carcinoma.

@@

Recently, there has been promising work on the bacterial microbiome and it's ability to complement existing screening methods such as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014]. There has also been research into how the bacterial microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].  A few studies have also shown how this microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the pathogenesis of CRC.  These studies have helped to provide a tantalizing link between the bacterial microbiome and CRC. Although these studies suggest that the bacterial microbiome might change after treatment there remains limited information on the bacterial microbiome before and after surgical removal of lesion (adenoma or carcinoma) and whether the community changes at all.

In this study we tested the hypothesis that the bacterial microbiome changes between pre (initial) and post (follow up) samples after removal of a lesion (adenoma or carcinoma).  Our analysis included both alpha and beta diversity analysis along with investigation of individual operational taxonomic units (OTUs).  We also utilized Random Forest to build models to classify either initial and follow up samples or lesion and normal samples and then subsequently observe how these models, as well as specific OTUs within them, performed pre and post surgery.  We also investigated the models for similar important OTUs to identify the crucial OTUs for not only classifying initial and follow up samples but also lesion or normal.

@@

***Cancer Associated Bacteria:***
Previous literature has suggested that a number of oral microbes may be important in CRC pathogenesis [@flynn_metabolic_2016].  So we next examined whether there were differences in previously well described carcinoma associated OTUs.  These included the OTUs that aligned with *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).  There was a decrease in relative abundance in initial and follow up samples for lesion and carcinoma for *Parvimonas micra* (P-value `r ifelse(main_crc_bugs_summary["parv", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05), and *Porphyromonas asaccharolytica* (P-value `r ifelse(main_crc_bugs_summary["porp", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05). In contrast, there was no difference in relative abundance in any of these OTUs for those with adenoma [Figure 2]. We also observed that only a small percentage of those with adenoma or carcinoma were positive or had an appreciable relative abundance of any of these respective OTUs [Figure 2].

***The Lesion Model:***
We next wanted to identify if there were any common bacterial microbiome changes in individuals with adenoma or carcinoma versus normal control individuals.  We investigated this by creating a model to classify lesion versus normal based on the bacterial community and FIT measurements.  This model had an AUC range of `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.  The ROC curve for the final lesion model used was within the observed range of the 100 different test set AUC iterations [Figure 3a].  There were a total of `r length(lesion_top10_varsMDA$variable)` variables [Figure S2].  For this model FIT measurement resulted in the largest decrease in MDA while the OTU with the largest MDA was *`r gsub("_unclassified", "", lesion_imp_vars[as.character((filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) [Figure S2a].

If there were common OTUs that could separate adenoma and carcinoma from normal controls, we would expect to find a decrease in the positive probability of the follow up sample to be a lesion. This is what we observed for the lesion model [Figure 4] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "lesion")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).  When we separated individuals based on whether they had an adenoma or carcinoma there was only a decrease in positive probability for the carcinoma group (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "carcinoma_only")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001) and not for the adenoma group (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "all_adenoma")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).   We also observed that there were no significant differences in whether the models classified the samples as having lesion between the predicted and actual (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_lesion" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05).  The lesion model was able to correctly classify the one individual who still had a carcinoma on follow up [Figure 4].

***The Initial Sample Model:***
After building a model to classify based on lesion we wanted to then build a separate model specifically to be able to classify whether samples were initial or follow up samples based on the bacterial community and FIT measurements.  The initial sample model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.  The test set AUC range for this model performed better then the training set AUCs.  There was a marked decrease in the ROC curve for the final model used when compared to the 100 test set AUC iterations [Figure 3b].

If there were common OTUs that could separate initial from follow up sample regardless of whether the lesion was adenoma or carcinoma we would expect to find a decrease in the positive probability of the follow up sample to be an initial sample. This is what we observed for the initial sample model [Figure 4] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "lesion")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).

***Common OTUs to both Models:***
The reduced models were built based on the most important variables to either classification model.  whereas for the initial sample model there were a total of `r length(IF_top10_varsMDA$variable)` variables [Figure S3].  For both models FIT measurement resulted in one of the largest decreases in MDA [Figure S2a & S3a].  When we compared the two different reduced models with each other there were a total of `r length(common_vars_summary$otu)` common OTUs.  Some of the most common taxonomic identifications belonged to *`r filter(common_vars_summary, otu == "Otu000028")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000018")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000009")[, "lowest_ID"]`*, and *`r filter(common_vars_summary, otu == "Otu000005")[, "lowest_ID"]`*.  The vast majority of these OTUs had classifications to bacteria typically thought of as commensal [Table S2].

@@

## Discussion
From our results there were large observed differences in the bacterial microbiome between pre and post surgery samples based on whether the individual had an adenoma or carcinoma.  There were much larger differences between initial and follow up samples based on the thetayc distance metric and in fecal blood as measured by FIT for individuals with carcinoma versus adenoma [Figure 1].  However, there were no differences between initial and follow up samples for Shannon Diversity, observed OTUs, or evenness regardless of whether the individual had an adenoma or carcinoma [Table S1].  There was also no differences in relative abundance of any specific OTU for lesion, adenoma only, or carcinoma only [Figure s1].  Although there was a detectable change in the carcinoma group.  It should be noted that whether this change was towards what would be be expected for normal controls is not known.

@@

***Analysis Overview:***
We first wanted to test if there were any differences based on whether the individual had an adenoma or carcinoma.  This was done by testing initial and follow up samples for differences in alpha and beta diversity, testing differences in FIT between initial and follow ups, testing all OTUs that were used in either built model, and investigating the relative abundance of specific previously associated CRC bacteria (*Fusobacterium nucleatum*, *Parvimonas micra*, *Peptostreptococcus assacharolytica*, and *Porphyromonas stomatis*) based on adenoma and carcinoma.  From here the lesion model was then tested for accuracy in prediction and whether it reduced the positive probability of lesion after surgery.  The most important OTUs for this were used to build a reduced model and it was assessed for similarity to the original model.  We then used the initial sample model to assess whether it could classify samples better then the lesion model.  The most important OTUs were then identified from this model and used to create a reduced feature initial sample model.  This reduced feature model, as was done with the lesion model, was compared to the full model for loss of accuracy.  A list of common OTUs were found for the two different models used.  Finally, both lesion and initial sample models initial positive probabilities were tested against the follow up positive probabilities.

@@

**Figure 4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Initial Sample Reduced Variable Models**  A) Lesion positive probability change from initial to follow up sample in those with carcinoma   B) Initial positive probability change from initial to follow up sample in those with carcinoma   C) Lesion positive probability change from initial to follow up sample if those with adenoma or Screen Relevant Neoplasia (SRN).    D) Initial positive probability change from initial to follow up sample in those with adenoma or SRN.  The dotted line represents the threshold used to make the decision of whether a sample was positive or not.

@@

**Figure S4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Before Sample Full Variable Models** A) Lesion positive probability change from initial to follow up sample in those with carcinoma   B) Initial positive probability change from initial to follow up sample in those with carcinoma   C) Lesion positive probability change from initial to follow up sample in those with adenoma or Screen Relevant Neoplasia (SRN).    D) Initial positive probability change from initial to follow up sample in those with adenoma or SRN.  The dotted line represents the threshold used to make the decision of whether a sample was positive or not.
